---
title: "HW3"
author: Hanyu Lu
date: October 2020
output: 
  pdf_document:
    latex_engine : "xelatex"
---

# Problem 1

```{r setup}
library(tidyverse)
library(arsenal)
library(ggplot2)
library(ggridges)
set.seed(996)
```


```{r echo = FALSE}
exercise_df = read_csv("./Exercise.csv")
```

```{r echo = FALSE}


exercise_df2 = mutate(exercise_df, Group2 = Group) %>% 
  mutate(
    Group = as.factor(Group),
    Group2 = as.factor(Group2),
  )

levels(exercise_df2$Group2) = 
  list(Systolic_POST_intervention = "1", Systolic_POST_control = "0")

levels(exercise_df2$Group) = 
  list(Systolic_PRE_intervention = "1", Systolic_PRE_control = "0")

exercise_df2  =
exercise_df2  %>%
pivot_wider(
  names_from = Group,
  values_from = Systolic_PRE, 
  ) %>% 
  pivot_wider(
  names_from = Group2,
  values_from = Systolic_POST, 
  )  %>% 
  mutate(
    control_difference = Systolic_POST_control - Systolic_PRE_control,
    intervention_difference = Systolic_POST_intervention - Systolic_PRE_intervention
  )

my_controls <- tableby.control(
                total = F,
                test = F,  # No test p-values yet
                numeric.stats = c("meansd", "medianq1q3"),
                stats.labels = 
                list(
                meansd = "Mean (SD)",
                medianq1q3 = "Median (Q1, Q3)")
                )

tab2 <- tableby( ~ Systolic_POST_intervention + Systolic_POST_control  + Systolic_PRE_intervention + Systolic_PRE_control + control_difference + intervention_difference, data=exercise_df2, control=my_controls)

summary(tab2, title = "Descriptive Statistics", text = T, digits = 2)
```

a) Perform appropriate tests to assess if the Systolic BP at 6 months is significantly different from the baseline values for each of the groups:
i) Intervention group (5p)

Since we don't know true population variance.  We are going to use paired t-test because we intend to compare scores on two different variables but on the same group. Additionally, we test for the mean of the differences with unknown variance.

$H_0$: the Systolic BP at 6 months is equal to the baseline values for intervention group
$H_1$: the Systolic BP at 6 months is significantly different from the baseline values for intervention group

$\bar{d} = \sum\limits_{i=1}^{n}d_i/n$ = -8.58
$s_d = \sqrt{\sum\nolimits_{i=1}^{n}(d_i-\bar{d})^2/(n-1)}$ = 17.17

$t = \frac{\bar{d}-0}{s_d/\sqrt{n}} = \frac{-8.58 - 0}{17.17/\sqrt{36}}$ = -3 ~ $t_{36-1}$

$t_{36-1,0.975}$ = 2.03

Since this t-test is two-sided, |t| = 3 > $t_{36-1,0.975}$ = 2.03.

We can reject $H_0$. We can conclude that the Systolic BP at 6 months is significantly different from the baseline values for intervention group

```{r echo = FALSE}
#intervention group

sd(exercise_df2$intervention_difference, na.rm = TRUE)

-8.58/(17.17/6)

qt(0.975,35) #t-test
```

ii) Control group (5p)

Since we don't know true population variance.  We are going to use paired t-test because we intend to compare scores on two different variables but on the same group. Additionally, we test for the mean of the differences with unknown variance.

$H_0$: the Systolic BP at 6 months is equal to the baseline values for control group
$H_1$: the Systolic BP at 6 months is significantly different from the baseline values for control group

$\bar{d} = \sum\limits_{i=1}^{n}d_i/n$ = -3.33
$s_d = \sqrt{\sum\nolimits_{i=1}^{n}(d_i-\bar{d})^2/(n-1)}$ = 14.81

$t = \frac{\bar{d}-0}{s_d/\sqrt{n}} = \frac{-3.33 - 0}{14.81/\sqrt{36}}$ = -1.35 ~ $t_{36-1}$

$t_{36-1,0.975}$ = 2.03

Since this t-test is two-sided, |t| = 1.35 < $t_{36-1,0.975}$ = 2.03.

We cannot reject $H_0$. We can conclude that the Systolic BP at 6 months is not significantly different from the baseline values for intervention group

```{r echo = FALSE}
#control group

sd(exercise_df2$control_difference, na.rm = TRUE)

-3.33/(14.81/6)

#t test
qt(0.975,35)
```

b) Now perform a test and provide the 95% confidence interval to assess the Systolic BP absolute changes between the two groups.

First test the hypothesis: $H_0: \sigma_1^2 = \sigma_2^2$ vs $H_1: \sigma_1^2 \neq \sigma_2^2$

With significance level $\alpha$ pre-specified, compute the test statistic:

$F = \frac{s_1^2}{s_2^2}$ = $F = \frac{17.17^2}{14.81^2}$ =1.34 ~ $F_{36-1,36-1}$

$F_{36-1,36-1, 0.975}$ = 1.96

$F_{36-1,36-1, 0.025}$ = 0.51

$F_{36-1,36-1, 0.025}$ = 0.51 < F = 1.34 < $F_{36-1,36-1, 0.975}$ = 1.96

Therefore we are unable to reject $H_0$ that the variances are equal.

Thus we can conduct two-sample independent t-test with equal variances.

$H_0$ = The Systolic BP absolute changes between the two groups is equal to 0
$H_1$ = The Systolic BP absolute changes between the two groups is not equal to 0

$s = \sqrt{\frac{(n_1-1)s_1^2+(n_2-1)s_2^2}{n_1+n_2-2}}$ = $\sqrt{\frac{(36-1)17.17^2+(36-1)14.81^2}{36+36-2}}$ = 16.03

$t = \frac{\bar{X1}-\bar{X2}}{s\sqrt{\frac{1}{n1}+\frac{1}{n2}}}$ = $\frac{-8.58-(-3.33)}{16.03\sqrt{\frac{1}{36}+\frac{1}{36}}}$ = -1.39 ~ $t_{36+36-2}$

$t_{36+36-2, 0.975}$ = 1.99

|t| = 1.39 < $t_{36+36-2, 0.975}$ = 1.99, so we fail to reject $H_0$, which means that there is no significant Systolic BP absolute changes between the two groups.

95% confidence interval:
($\bar{X1} - \bar{X2} -t_{n_1+n_2-2,1-\alpha/2}*s*\sqrt{\frac{1}{n1}+\frac{1}{n2}}$, $\bar{X1} - \bar{X2} +t_{n_1+n_2-2,1-\alpha/2}*s*\sqrt{\frac{1}{n1}+\frac{1}{n2}}$)
= (-8.58-(-3.33) - 1.99 * 16.03 * $\sqrt{\frac{1}{36}+\frac{1}{36}}$, -8.58-(-3.33) + 1.99 * 16.03 * $\sqrt{\frac{1}{36}+\frac{1}{36}}$) = (-12.77, 2.27)

```{r}
# F test statistic
17.17^2/(14.81^2)

qf(0.975,35,35)

qf(0.025,35,35)

#two-sample independent t-test with equal variances

sqrt((35*17.17^2+35*14.81^2)/70) #s

(-8.58-(-3.33))/(16.03*sqrt(1/36+1/36)) #t

qt(0.975,70)

# Confidence Interval

-8.58-(-3.33) - 1.99 * 16.03 * sqrt(1/36+1/36)  #lower
-8.58-(-3.33) + 1.99 * 16.03 * sqrt(1/36+1/36)  #upper
```

c) What are the main underlying assumptions for the tests performed in parts a) and b)? 
i) Use graphical displays to check the normality assumption and discuss the findings.

```{r}
par(mfrow = c(1, 2))

hist(exercise_df2$control_difference, breaks=seq(-40,40,6),xlab="control difference", main ="control")


hist(exercise_df2$intervention_difference, breaks=seq(-40,40,6),xlab="intervention difference", main ="intervention")


```



# Problem 2

a) Generate one random sample of size n=20 from the underlying (null) true distribution. Calculate the test statistic, compare to the critical value and report the conclusion: 1, if you reject ð»0 or 0, if you fail to rejected ð»0. 

$H_0$ = The average IQ score of Ivy League colleges is equal to 120
$H_1$ = The average IQ score of Ivy League colleges is less than 120

$z = \frac{\bar{X}-\mu}{\sigma/\sqrt{n}}$ = -0.89

$z_{0.05}$ = -1.64

$z = -0.89 > z_{0.05} = -1.64$.  We fail to reject $H_0$.  We conclude that the average IQ score of Ivy League colleges is equal to 120.

```{r}
random_sample = rnorm(20, mean = 120, sd = 15)

z = (mean(random_sample)-120)/(sd(random_sample)/sqrt(20))

qnorm(0.05)

print(ifelse(z < -1.64, 1, 0))
```

By using the code ifelse(z < -1.64, 1, 0), the output is 1.

b) Now generate 100 random samples of size n = 20 from the underlying (null) true distribution and repeat the process in part (a) for each sample (calculate the test statistic, compare to the critical value, and record 1 or 0 based on criteria above). Report the percentage of 1s and 0s respectively across the 100 samples. The percentage of 1s represents the type I error.

```{r}

sample_means_rep1 = rep(NA, 100)
sample_z1 = rep(NA, 100)
random_sample_2 <- list(mode="vector",length=100)

for(i in 1:100){
  random_sample_2[[i]] = rnorm(20, mean = 120, sd = 15)
  #Generate 100 sample z-scores
  sample_z1[i] = (mean(random_sample_2[[i]])-120)/(sd(random_sample_2[[i]])/sqrt(20))
  sample_means_rep1[i] = ifelse(sample_z1[i] < -1.64, 1, 0)
  #Report 1 or 0
}

count1_1 =sum(sample_means_rep1==1)
count0_1 = sum(sample_means_rep1==0)

print(as.data.frame(table(sample_means_rep1)))

```

The percentage of 0's in the 100 samples is `r count0_1`% while the percentage of 1's in the 100 samples is `r count1_1`%.

c) Now generate 1000 random samples of size n = 20 from the underlying (null) true distribution, repeat the same process, and report the percentage of 1s and 0s across the 1000 samples.

```{r}
sample_means_rep2 = rep(NA, 1000)
sample_z2 = rep(NA, 1000)
random_sample_3 <- list(mode="vector",length=1000)

for(i in 1:1000){
  random_sample_3[[i]] = rnorm(20, mean = 120, sd = 15)
  #Generate 1000 sample z-scores
  sample_z2[i] = (mean(random_sample_3[[i]])-120)/(sd(random_sample_3[[i]])/sqrt(20))
  sample_means_rep2[i] = ifelse(sample_z2[i] < -1.64, 1, 0)
  #Report 1 or 0
}

count1_2 =sum(sample_means_rep2==1)
count0_2 = sum(sample_means_rep2==0)
print(as.data.frame(table(sample_means_rep2)))
```

The percentage of 0's in the 1000 samples is `r count0_2/10`% while the percentage of 1's in the 1000 samples is `r count1_2/10`%.

d) Final conclusions: compare the type I errors (percentage of 1s) from part b) and c). How do they compare to the level that we initially imposed (i.e. 0.05)? Comment on your findings.

Type I error, i.e., P(reject $H_0 \mid H_0$ is true) of 100 samples is 0.06, while that of 1000 samples is 0.055.  Since $\alpha$ level is 0.05, we may infer that as times of sample increase, the type I error of sampling distribution becomes closer to the significance level we imposed.