---
title: "HW3"
author: Hanyu Lu
date: October 2020
output: 
  pdf_document:
    latex_engine : "xelatex"
---

# Problem 1

```{r}
library(tidyverse)
library(arsenal)
```


```{r}
exercise_df = read_csv("./Exercise.csv")
```

```{r}


exercise_df2 = mutate(exercise_df, Group2 = Group) %>% 
  mutate(
    Group = as.factor(Group),
    Group2 = as.factor(Group2),
  )

levels(exercise_df2$Group2) = 
  list(Systolic_POST_intervention = "1", Systolic_POST_control = "0")

levels(exercise_df2$Group) = 
  list(Systolic_PRE_intervention = "1", Systolic_PRE_control = "0")

exercise_df2  =
exercise_df2  %>%
pivot_wider(
  names_from = Group,
  values_from = Systolic_PRE, 
  ) %>% 
  pivot_wider(
  names_from = Group2,
  values_from = Systolic_POST, 
  )  %>% 
  mutate(
    control_difference = Systolic_POST_control - Systolic_PRE_control,
    intervention_difference = Systolic_POST_intervention - Systolic_PRE_intervention
  )

my_controls <- tableby.control(
                total = F,
                test = F,  # No test p-values yet
                numeric.stats = c("meansd", "medianq1q3"),
                stats.labels = 
                list(
                meansd = "Mean (SD)",
                medianq1q3 = "Median (Q1, Q3)")
                )

tab2 <- tableby( ~ Systolic_POST_intervention + Systolic_POST_control  + Systolic_PRE_intervention + Systolic_PRE_control + control_difference + intervention_difference, data=exercise_df2, control=my_controls)

summary(tab2, title = "Descriptive Statistics", text = T, digits = 2)
```

a) Perform appropriate tests to assess if the Systolic BP at 6 months is significantly different from the baseline values for each of the groups:
i) Intervention group (5p)

Since we don't know true population variance.  We are going to use paired t-test because we intend to compare scores on two different variables but on the same group. Additionally, we test for the mean of the differences with unknown variance.

$H_0$: the Systolic BP at 6 months is equal to the baseline values for intervention group
$H_1$: the Systolic BP at 6 months is significantly different from the baseline values for intervention group

$\bar{d} = \sum\limits_{i=1}^{n}d_i/n$ = -8.58
$s_d = \sqrt{\sum\nolimits_{i=1}^{n}(d_i-\bar{d})^2/(n-1)}$ = 17.17

$t = \frac{\bar{d}-0}{s_d/\sqrt{n}} = \frac{-8.58 - 0}{17.17/\sqrt{36}}$ = -3

$t_{36-1,0.975}$ = 2.03

Since this t-test is two-sided, |t| = 3 > $t_{36-1,0.975}$ = 2.03.

We can reject $H_0$. We can conclude that the Systolic BP at 6 months is significantly different from the baseline values for intervention group

```{r}
sd(exercise_df2$intervention_difference, na.rm = TRUE)

-8.58/(17.17/6)

qt(0.975,35)
```

ii) Control group (5p)

Since we don't know true population variance.  We are going to use paired t-test because we intend to compare scores on two different variables but on the same group. Additionally, we test for the mean of the differences with unknown variance.

$H_0$: the Systolic BP at 6 months is equal to the baseline values for control group
$H_1$: the Systolic BP at 6 months is significantly different from the baseline values for control group

$\bar{d} = \sum\limits_{i=1}^{n}d_i/n$ = -3.33
$s_d = \sqrt{\sum\nolimits_{i=1}^{n}(d_i-\bar{d})^2/(n-1)}$ = 14.81

$t = \frac{\bar{d}-0}{s_d/\sqrt{n}} = \frac{-3.33 - 0}{14.81/\sqrt{36}}$ = -1.35

$t_{36-1,0.975}$ = 2.03

Since this t-test is two-sided, |t| = 1.35 < $t_{36-1,0.975}$ = 2.03.

We cannot reject $H_0$. We can conclude that the Systolic BP at 6 months is not significantly different from the baseline values for intervention group

```{r}
sd(exercise_df2$control_difference, na.rm = TRUE)

-3.33/(14.81/6)

qt(0.975,35)
```

# Problem 2

a) Generate one random sample of size n=20 from the underlying (null) true distribution. Calculate the test statistic, compare to the critical value and report the conclusion: 1, if you reject ð»0 or 0, if you fail to rejected ð»0. 

```{r}
random_sample = rnorm(20, mean = 120, sd = 15)
```

